Combining Language Resources Into A Grammar-Driven Swedish Parser

### 2012-03-05 doc/paper/final/PaperFin.tex
third final lrec

\documentclass[10pt, a4paper]{article}
\usepackage{lrec2006}
\usepackage{lingmacros}
\usepackage{graphicx}
%\usepackage[utf8]{inputenc}

\title{Combining Language Resources Into A Grammar-Driven Swedish parser}

@@

\section{Introduction}
% Copied from Master's Thesis. ok to do that?
Our goal is to implement a wide-coverage grammar and parser for Swedish
using the GF formalism, and 
thereby investigate how GF can be used for open-domain parsing.
We compile a large-scale grammar to a parser
and combine it with the extensive lexicon SALDO. The Swedish treebank 
Talbanken provides manually tagged trees which we use for improving and evaluating
the grammar. The parser will additionally
be evaluated by experts. \\
We chose our resources so that both our grammar and parser can 
be freely available and open-source. 
The target language is Swedish, a North-Germanic language
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
one of the official languages in Finland and altogether spoken by approximately 9
million people.
Swedish syntax is often similar to English, but the  morphology is richer and the
word order slightly more intricate.
It is a verb-second language: the second constituent of a declarative main
clause must consist of a verb.
The first constituent of the clause is usually made up of the subject,
although it likewise could consist of adverbial phrases or objects.
Fronting the finite verb marks questions.

@@

\subsection{SALDO}
SALDO \cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon. It is
developed at Spr{\aa}kbanken at Gothenburg University
and intended for usage in language technology
research. 
We have developde tools for extracting GF lexicons
from SALDO, described in \ref{sec:extractsaldo}
%which have resulted in a dictionary containing
%over 98000 entries including information about verbs that
%are reflexive or needs a particle.
%The importing method should be fast and reliable
%enough to allow us to always have a fresh version of the dictionary
%in GF.

\section{Related work}
%Copied form thesis!!!! 
% Add: how are these related to this work?
% Add :detaild comparison of outputs of various parsers of Sw.
Many years of research have lead to many interesting language
technology tools for Swedish.
An example is the well-known data-driven Malt parser \cite{malt},
trained on Talbanken. 
There are also a number of grammar-based parsers, although none is freely available.
The cascaded finite state parser CassSwe \cite{casswe} and
The Swedish Constraint Grammar \cite{birn}
give syntactic analyses. 
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
\cite{fdg},
an extension of the Constraint Grammar
formalism, and produces a dependency structure focusing on finding the nominal
arguments. \\

@@

%old related work:
%Language technology for Swedish is an area of much interesting research.
%There are two other deep grammar-based parser for Swedish:
% put in references!!
%Swedish Core Language Engine (Raynen and Gambäck, 1992) and a
% does this really exist??
%unification-based chart parser \cite{wiren},
%though none of them are freely available.
%Among other parsers for Swedish, the statistical MaltParser \cite{malt},
%trained on Talbanken, is worth mentioning. 
% put in references!!
%CassSwe (Kokkinakis and Johansson-Kokkinakis, 1999) is based on finite state cascades,
% put in references!!
%whereas the shallow parser GTA (Knutsson, Bigert and Kann, 2003) relies on rules of 
%a context free grammar. Both GTA and CassSwe operate on POS tagged text.\\
%Extract and FM \cite{MarkusForsberg2007} are tools for supervised lexicon
%extraction compatible with GF.

%\section{Results and progress}
%\label{sec:progress}

@@

\subsubsection{Results}
\label{sec:saldoRes}
The resulting dictionary contains more than 100 000 entries, approximately 80 \% 
of the total size of SALDO.
There are a number of reasons why some words were not imported,
the most obvious one is that we do not want all categories from
SALDO in the GF lexicon. Prepositions, 
numerals, 
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
SALDO contains many pronouns which
are not analysed the same way in GF. %TODO see where??
Before adding them to our lexicon, we need to do more analysing to find their
correct GF-category. 
%Some experiments on finding the category
%have been done using Talbanken, see section \ref{sec:gf.quant}.
Categories involving multiple words 
are usually handled as idioms and should be given in a separate lexicon. In
total six types of words were considered for the extraction: \\

@@

%Since SALDO does not give any valency information, neither does the imported lexicon.
%The lexicon does not contain any information.
\noindent Valency information, which is crucial for GF, is not given in SALDO and
hence not in the imported lexicon. Instead we extract this from
Lexin\footnote{http://spraakbanken.gu.se/lexin/}.

\subsection{A tool for lexical acquisition}
As we extract our main lexicon from SALDO, we have also created a tool for
semi-automatic acquisition to complement the lexicon.
%We have tested it on verbs, with good results.
Like the SALDO importer it makes use of
the smart paradigm given in the Swedish resource grammar.
Unlike the SALDO importer, this tool does not require any particular
word form, but operates on any verb form and 
%By combining this method with the information from the tags in Talbanken,
%the tool interactively generates GF lexical entries. 
%Given a list of words, it 
iteratively tries to figure out how to conjugate each of them. If several forms
of a word are 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection. 
If the table contains all other conjugations from the input list,
the program will ask the  user to
validate the claimed paradigm. 
This step is needed since the input may not provide information enough
to automatically do the validation. The user may now either
allow the word to be added to the lexicon, remove it or request another guess.
The user hence only needs to decide if each paradigm is correct or not, and
the program will do the rest.

@@

\section{Translating The Treebank Talbanken}
\label{sec:mapping}
The information from the tags in Talbanken can be used for many purposes.
We have developed an automatic transformation of Talbanken trees 
to trees in GF format. The translation makes use of the POS tags as well as
the syntactic information. 
\begin{figure}[h!]
\begin{center}
\hspace{-30mm}
\subfloat{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\hspace{-10mm}
\subfloat{\label{pic:gftree}\includegraphics[width=90mm]{gfTree.pdf}}
%\subfloat{\label{pic:gfptree}\includegraphics[width=50mm]{gfPTree.pdf}}
\caption{The Talbanken tree and the GF tree for the sentence ``Katten p{\aa} bilen blir st{\"o}rre".}
\label{fig:translationtrees}
\end{center}
\end{figure}

@@

The translation %connects GF with another annotations and 
gives us means to evaluate our parser. By both parsing a Talbanken sentence and
transforming its annotated tree, we can easily inspect if the results are
equal.
Additionally, the mapping shows which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities of how often a function is
used. The GF treebank we achieve from the translation is a good source for this
information.\\

%The mapping gives information about which form a word is
%currently used in and this may be used by the lexical extraction
%tools.
%The translated trees enable us to extract probabilities for how often
%different functions are used, a feature that will enable disambiguation.
%Furthermore, the translation makes it easy to identify grammatical constructions
%missing from the GF grammar and shows how the GF analysis differs from the one made
%in Talbanken.
%Another important use of the mapping is evaluation of the parser, which can be
%accomplished by comparing the parse trees and the trees from the transformer.
%\subsection{Differences in the notation}
%Some structural differences between GF and Talbanken became obvious during the
%translation.
%In Talbanken the valency information is given implicitly by the complements
%of a word. If a verb is followed by an object, \verb-OO-, which contains an
%\verb-S-, we can conclude that this is a sentence-complement verb.
%In GF, the valency information is given for each entry in the lexicon.
%A sentence-complement verb has the type \verb-VS- and from this we know 
%that the function \verb-ComplVS- should be used to combine the verb and with a 
%sentence.
%\begin{wrapfigure}{l}{0.52\textwidth}
%\includegraphics[width=80mm]{metatree.pdf}
%\caption{Abstract tree with meta variable}
%\label{pic:gMetatree}
%\end{wrapfigure}
%
%\noindent A similar issue arises from the use of verb particles. A verb with a
%particle is treated as two independent units in Talbanken but in GF they are
%considered to be one constituent. % which means that 
%The particle has therefore no effect on the abstract tree, its presence is
%announced by the verb \\itself.
%
%
%We have so far concentrated on shorter sentences, without idioms and
%conjunction. 
%Since GF demands information 
%that is not given by the Talbanken tags -- such as the valency of verbs -- a
%full translation requires that all words used are already known to GF. 
%The evaluation of trees of this sort shows that the mapping
%can restore 85\% of the nodes. 
%If we lift the restrictions, we can restore 63\% of the tags.
\subsection{Results and evaluation}
%>> more! replace words, how much better results (both grammar and mapping) 
%>> more results
%>> motivation, why so bad??\\
The development was mostly example-driven, and at least one rule for the translation of
every Talbanken-tag has been implemented.
Shorter sentences have been prioritized in order to get a coverage of the most
fundamental constructions. 
Regression testing was used to ensure that a new rule would not violate the old ones
or allow trees to be translated erroneously.

The flat version of Talbanken05 was used for developing the mapping, but
the program works for both the flat and the deep Tiger-XML versions.
%The deeper contains six more tags, although they are not always useful for our       
%purpose. Figure \ref{fig:mappDeepFlat} shows the difference for a conjunction.
%% show piece of s5 here!
%\begin{figure}[h]
%\begin{parsetree}
%    ( .CNP. 
%        (.CJ.  (.NP. ~ `oskifta dödsbon'))
%        ( .++. (.++OC. `och'))
%        (.CJ.  (.VN. `familjestiftelser'))
%    )
%\end{parsetree}
%\begin{parsetree}
%    ( .CNP. 
%        (.CJ.  (.NP. ~ `oskifta dödsbon'))
%        ( .CC.
%            (.CNP. 
%                (.++.  (.++OC. `och' ))
%                (.C+.  (.VN. `familjestiftelser'))
%         ))
%    )
%\end{parsetree}
%\caption{An example of the difference in the two analyses}
%\label{fig:mappDeepFlat}
%\end{figure}
The focus on the mapping has been simple sentences, in order to cover the most
fundamental structures. 
%If the program is extended and adapted to cover complex
%conjunction, a deeper evaluation should be preformed to see which version that
%suits our needs the best.
%The deeper implementation also gives 
%some more information about verb phrases, by the tag \verb|VG| (\emph{verb group}),
%which groups an infinitival verb and its object together into a \verb|VP|.
%This information can however be extracted from the flat implementation, and the results
%get slightly better when using this version. \\

When evaluating the mapping, the results strongly depend on which restrictions we
put on the input. 
One of the reasons why a node cannot be translated, is the 
use of the tags show in figure \ref{fig:mapBadtag}.
The \textbf{PU} tag is used for graphic listings, and not for fluent text.
In our grammar there is naturally no corresponding function; 
%We are not interested in sentences made up by list, since 
the listings are meant for making the text look nice in
folders etc and outside the scope for the grammar itself. The tags \textbf{XX}
and \textbf{NAC} are often used since Talbanken makes a
difference between subject
and object noun phrases. %(which is not needed in GF (see section \ref{sec:swegf}))
The analysis of elliptical expression in (\ref{sent:krav})
\enumsentence{F\"or stora krav.\\
``Too high demands."}\label{sent:krav}
contains the tags \textbf{XX} and \textbf{NAC}, since it is not obvious
whether the noun phrase is used as subject or an object.
The tags shown in figure \ref{fig:mapBadtag} occur quite frequently in the treebank and are always translated
to metas, which lowers our result. \\
%How we cannot cover all cases. 
\begin{figure}[h]
\begin{tabular}{ll}
\textbf{NAC} & Not a constituent\\
\textbf{XP} & Other (non-coordinated) phrase\\
\textbf{DB} & Doubled function\\
\textbf{PU} & List item\\
\textbf{XX} & Unclassifiable part-of-speech\\
\end{tabular}
\caption{}\label{fig:mapBadtag}
\end{figure}

The main goal has been to be able to translate shorter sentences, with no
idioms or conjunction.
If we assure that the lexicon contains the correct word class for all lemmas
involved, we
can restore more than 85 \% of the nodes in the original tree.
If we lift all the restrictions excluding the \verb|PU|, we get
%When tested the 100 first sentences from Talbanken, we get
65 \% coverage. 
If we test randomly collected sentences that do not contain any of the tags listed
in figure \ref{fig:mapBadtag}, 72 \% can be restored (see figure \ref{tab:mappres})

%TODO
%All NPs: 72%
\begin{figure}[h]
\begin{tabular}{|ll|}
\hline
No list items & 65 \%\\
No special punctuation or bad tags& 72 \%\\
Short sentences with known words & 85 \%\\
\hline
\end{tabular}
\caption{}
\label{tab:mappres}
\end{figure}

A mapping between GF and the Wall Street Journal of Penn Treebank has earlier been conducted \cite{gfMech}.
The percentage of restored nodes from Peen Treebank is higher than our results.
%TODO review
The reason for this may be the fact that English is syntactically less complicated
than Swedish.
Furthermore, the text in Talbanken are from various brochures, newspapers and text books,
where idiomatic expressions are more likely the be appear and the language
presumably less strict than in Wall Street Journal\footnote{www.wsj.com}.
Also, the Penn Treebank contains a lower number of tags, 82 compared
to more than 300 in Talbanken. Even if the tags describing the same word
%TODO review
class as another tag are excluded, Talbanken still leaves us with more than 130 tags.
With more tags, we get more information, but as the number 
%Thus the number of
increase, so does the amount of work of finding the correct translation 
for each combination of tags and writing rules that cover all constructions. 
%iand so is the number of
%rules that would be needed to cover all constructions.

@@

It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no large resource available to evaluate it on.
Our evaluations are based on Talbanken, and when first conducting tests,
we found much room for improvement.
%For Swedish, about 85\% of the GF resource code is shared with the other Scandinavian
%languages. 
%However, if we aim for a deeper and more comprehensive analysis of Swedish,
the implementation of the languages needs to be more independent.
The resource grammar gives a good start and our present grammar 
% don't say inverted word order
covers constructions such as declarative sentences, questions, passives,
imperatives, relative clauses, cleft constructions etc.
During our work the grammar has been extended to cover language specific
constructions such as topicalisation and the features listed below. We further
put effort on avoiding and removing overgeneration, which might pass unnoticed
when parsing controlled text but which cause an explosion of ambiguous parse trees
when used for free text and with a large lexicon.
\subsection{The s-passive}
Passive voice is commonly used in Swedish.
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using the modal auxiliary verb `bli'
(\emph{`become'}) and the \textbf{s-passive} which is formed by adding an
\emph{s} to the verb: \\
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {\emph{the essay} & \emph{wrote+\textbf{s}} & \emph{by} & \emph{a} & \emph{student}}
        {``The essay was written by a student."}}\label{sent:skrevs}
%{b. &En & student & skrev & uppsatsen.}
%        {&a &student &wrote &the essay}
%        {~~`A student wrote the essay.'}}
Some studies suggest that the s-passive is used in more than 80 \% of the times
\cite{laanemets}.
It is however not as common in the other Scandinavian languages,
and the resource grammar for Scandinavian therefore only implemented a function for
passive by using auxiliary verb.

@@

\enumsentence{Han s{\aa}g sina barns skor.\\ \label{ex:ReflGen}
He saw {\sc self's} children's shoes.}
\enumsentence{Sina vantar hade han gl{\"o}mt p{\aa} t{\aa}get.\\
{\sc self's} gloves, he had forgotten on the train.}
\enumsentence{Hon ber alla sina kompisar att g{\aa}\\
She asks all {\sc self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
%\enumsentence{\begin{tabular}[t]{@{}*{5}{l@{\ }}} 
%a. & Han är längre än sin kompis. & & b. &Han är här oftare än sin kompis.\\
%   & He is taller than {\sc self's} friend.
%   & & & He is here more often than {\sc self's} friend.
%\end{tabular}}\label{ex:ReflAP} \label{ex:ReflAdv}
\eenumsentence{
\item Hon tyckte om skolan och alla sina elever.\\
     She liked the school and all {\sc self's} students.
\item Han s{\aa}g sina f{\aa} b{\"o}cker och sin penna.\\
 He saw {\sc self's} few books and {\sc self's} pencil.
}\label{ex:ReflConj} \label{ex:ReflConj2}
Reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in 
sentence (\ref{sent:sesig}) and (\ref{ex:ReflSubj}).
The third person reflexives (`sig',`sin') requires a third person antecedent (see \ref{ex:refljag}).
Furthermore, the antecedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).
The grammar does not accept any of these sentences:
\enumsentence{
*Sina vantar var kvar p{\aa} t{\aa}get.\\
{\sc self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Han och sin kompis läser en bok.\\
He and {\sc self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give {\sc self's} money to {\sc self's} siblings. }\label{ex:refljag}
\enumsentence{*Han vill att jag ser p{\aa} sig.\\
He wants me to look at {\sc self}.}\label{sent:sesig}

In order to achieve a scalable implementation of reflexive pronouns, we needed
to diverge from the old abstract implementation. To capture the agreement
between the antecedent and the reflexive pronoun, while still allowing the grammar
to accept reflexive constructions embedded in abverbial and adjectival phrases,
in fronted positions etc. we make use of the dependent types of GF. This is the
first use of the dependent types in the resource grammars and a could be
extended to other languages having similar dependecies.

\subsection{A second future tense: ``Kommer att"}
Swedish use two auxiliary verbs for future tense \\
\cite[p. 246]{H&H}:
%\cite[Tempus \textsection 28]{SAG}.
\eenumsentence{
\item {
Det \textbf{kommer att} bli m{\"o}rkt snart\\
``It will get dark soon" }
\item {
Jag \textbf{ska} g{\aa} och l{\"a}gga mig nu\\
``I'm going to bed now"\\}
}
The resource grammar included 
\emph{`ska'}, which was implemented as the standard way of forming future
tense, and hence represents the translation of
\emph{`will'}. 
The new grammar also supports \emph{``kommer att"}, expressed as an alternative
future tense.
\begin{figure}[h]
\begin{verbatim}
(UseCl (TTAnt TFutKommer ASimul) PPos 
       (ImpersCl (AdvVP (ComplVA become_VA 
                (PositA dark_A)) soon_Adv))
\end{verbatim}
\caption{Result of parsing ``Det kommer att bli m{\"o}rkt snart". The future tense
         is marked by the constant TFutKommer}
  \label{fig:kommeratt}
\end{figure}

@@

Parsing bigger parts of Talbanken would hence give very low results at this stage, 
%, for the reasons stated.
and a comparison of the results would not be of much value as
there would not be enough material %underlag is yet too small 
to do be able to do any interesting analysis.

@@

%\section{Evaluation and future work}
%We are building a parser which is based on grammatical rules rather than
%statistics. The parser will therefore validate that
%a given sentence is grammatically correct according to the rules defined.
%Chosen parts of the parser or grammar may be used in other applications
%which deal with controlled natural languages. Automatic language
%generation from the grammar is also provided by GF.
%
%%TODO already done?
%We aim to make the parser robust by equipping it with techniques such as chunk
%parsing, named entity recognition and methods for handling unknown 
%grammatical constructions such as idioms and ellipses. 
%It is already possible to add probabilities for GF functions,
%which rank the parse trees.
%By adding dependency probabilities, we aim to improve the
%potential for disambiguation.

%Use 10 \% of Talbanken as evaluation material, not seen by grammar developer.
The result will be evaluated both automatically -- by comparing the output of
the translated trees from Talbanken -- and manually by professor Elisabet Engdahl.
She also evaluates the intermediate results.

@@

%By combining GF with a large lexicon, we acquire a deep parser for 
%Swedish. All parts, including the underlying grammar, are open-source and can be re-used.
%We have developed tools for extending the lexicon with words from Talbanken,
%and we show how to make use of the information in the manually tagged
%treebank.
The usage of GF allows us to start from
a well-defined system for describing grammar, as well as tools for
parsing.

>>>>

\documentclass[10pt, a4paper]{article}
\usepackage{lrec2006}
\usepackage{lingmacros}
\usepackage{graphicx}

\title{Combining Language Resources Into A Grammar-Driven Swedish Parser}

@@

% TODO add links to the website?
\section{Introduction}
% Copied from Master's Thesis. ok to do that?
Our goal is to implement a wide-coverage grammar and parser for Swedish
using the GF formalism, and 
thereby investigate how GF can be used for open-domain parsing.
We compile a large-scale grammar to a parser
and combine it with the extensive lexicon SALDO. The Swedish treebank 
Talbanken provides manually tagged trees which we use for improving and evaluating
the grammar. The parser will additionally
be evaluated by experts. \\
We chose our resources so that both our grammar and parser can 
be freely available and open-source. 
The target language is Swedish, a North-Germanic language
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutually intelligible. Swedish is also 
one of the official languages in Finland and altogether spoken by approximately 9
million people.
Swedish syntax is often similar to English, but the  morphology is richer and the
word order slightly more intricate.
It is a verb-second language: the second constituent of a declarative main
clause must consist of a verb.
The first constituent of the clause is usually made up of the subject,
although it likewise could consist of adverbial phrases or objects.
Fronting the finite verb marks questions.

@@

\subsection{SALDO}
SALDO \cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon. It is
developed at Spr{\aa}kbanken at Gothenburg University
and intended for usage in language technology
research. 
We have developed tools for extracting GF lexicons
from SALDO, described in \ref{sec:extractsaldo}

\section{Related work}
%Copied form thesis!!!! 
% Add: how are these related to this work?
% Add: detaild comparison of outputs of various parsers of Sw.
Many years of research have lead to many interesting language
technology tools for Swedish.
An example is the well-known data-driven Malt parser \cite{malt},
trained on Talbanken. 
There are also a number of grammar-based parsers, although none is freely available.
The cascaded finite state parser CassSwe \cite{casswe} and
The Swedish Constraint Grammar \cite{birn}
give syntactic analyses. 
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
\cite{fdg},
an extension of the Constraint Grammar
formalism, and produces a dependency structure focusing on finding the nominal
arguments. \\

@@

@@

\subsubsection{Results}
\label{sec:saldoRes}
The resulting dictionary contains more than 100 000 entries, approximately 80 \% 
of the total size of SALDO.
There are a number of reasons why some words were not imported,
the most obvious one is that we do not want all categories from
SALDO in the GF lexicon. Prepositions, 
numerals, 
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
SALDO contains many pronouns which
are not analyzed the same way in GF. 
Before adding them to our lexicon, we need to do more analyzing to find their
correct GF-category. 
Categories involving multiple words 
are usually handled as idioms and should be given in a separate lexicon. In
total six types of words were considered for the extraction: \\

@@

Valency information, which is crucial for GF, is not given in SALDO and
hence not in the imported lexicon. Instead we extract this from
Lexin\footnote{http://spraakbanken.gu.se/lexin/}.

\subsection{A tool for lexical acquisition}
As we extract our main lexicon from SALDO, we have also created a tool for
semi-automatic acquisition to complement the lexicon.
Like the SALDO importer it makes use of
the smart paradigm given in the Swedish resource grammar.
Unlike the SALDO importer, this tool does not require any particular
word form, but operates on any verb form and 
iteratively tries to figure out how to conjugate each of them. If several forms
of a word are 
given, the program will try to identify the one that carries the most linguistic
information, put this in a form recognized by the smart paradigm and ask GF to output
a table with the resulting inflection. 
If the table contains all other conjugations from the input list,
the program will ask the  user to
validate the claimed paradigm. 
This step is needed since the input may not provide information enough
to automatically do the validation. The user may now either
allow the word to be added to the lexicon, remove it or request another guess.
The user hence only needs to decide if each paradigm is correct or not, and
the program will do the rest.

@@

\section{Translating The Treebank Talbanken}
\label{sec:mapping}
The information from the tags in Talbanken can be used for many purposes.
We have developed an automatic transformation of Talbanken trees 
to trees in GF format. The translation makes use of the POS tags as well as
the syntactic information. 
\begin{figure}[h!]
\begin{center}
\hspace{-30mm}
\subfloat{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\hspace{-10mm}
\subfloat{\label{pic:gftree}\includegraphics[width=90mm]{gfTree.pdf}}
\caption{The Talbanken tree and the GF tree for the sentence ``Katten p{\aa} bilen blir st{\"o}rre".}
\label{fig:translationtrees}
\end{center}
\end{figure}

@@

The translation 
gives us means to evaluate our parser. By both parsing a Talbanken sentence and
transforming its annotated tree, we can easily inspect if the results are
equal.
Additionally, the mapping shows which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities of how often a function is
used. The GF treebank we achieve from the translation is a good source for this
information.\\

\subsection{Results and evaluation}
The development was mostly example-driven, and at least one rule for the translation of
every Talbanken-tag has been implemented.
Shorter sentences have been prioritized in order to get a coverage of the most
fundamental constructions. 
Regression testing was used to ensure that a new rule would not violate the old ones
or allow trees to be translated erroneously.

The flat version of Talbanken05 was used for developing the mapping, but
the program works for both the flat and the deep Tiger-XML versions.
The focus on the mapping has been simple sentences, in order to cover the most
fundamental structures.

When evaluating the mapping, the results strongly depend on which restrictions we
put on the input. 
One of the reasons why a node cannot be translated, is the 
use of the tags show in figure \ref{fig:mapBadtag}.
The \textbf{PU} tag is used for graphic listings, and not for fluent text.
In our grammar there is naturally no corresponding function; 
the listings are meant for making the text look nice in
folders etc and outside the scope for the grammar itself. The tags \textbf{XX}
and \textbf{NAC} are often used since Talbanken makes a
difference between subject
and object noun phrases.
The analysis of elliptical expression in (\ref{sent:krav})
\enumsentence{F\"or stora krav.\\
``Too high demands."}\label{sent:krav}
contains the tags \textbf{XX} and \textbf{NAC}, since it is not obvious
whether the noun phrase is used as subject or an object.
The tags shown in figure \ref{fig:mapBadtag} occur quite frequently in the treebank and are always translated
to metas, which lowers our result. \\
\begin{figure}[h]
\begin{tabular}{ll}
\textbf{NAC} & Not a constituent\\
\textbf{XP} & Other (non-coordinated) phrase\\
\textbf{DB} & Doubled function\\
\textbf{PU} & List item\\
\textbf{XX} & Unclassifiable part-of-speech\\
\end{tabular}
\caption{Untranslatable tags}\label{fig:mapBadtag}
\end{figure}

The main goal has been to be able to translate shorter sentences, with no
idioms or conjunction.
If we assure that the lexicon contains the correct word class for all lemmas
involved, we
can restore more than 85 \% of the nodes in the original tree.
If we lift all the restrictions excluding the \verb|PU|, we get
65 \% coverage. 
If we test randomly collected sentences that do not contain any of the tags listed
in figure \ref{fig:mapBadtag}, 72 \% can be restored (see figure \ref{tab:mappres})

\begin{figure}[h]
\begin{tabular}{|ll|}
\hline
No list items & 65 \%\\
No special punctuation or bad tags& 72 \%\\
Short sentences with known words & 85 \%\\
\hline
\end{tabular}
%TODO is this correct???
\caption{Number of nodes in each translated tree not put to meta}
\label{tab:mappres}
\end{figure}

A mapping between GF and the Wall Street Journal of Penn Treebank has earlier
been conducted \cite{gfMech}.
The percentage of restored nodes from Peen Treebank is higher than our results.
The reason for this may be the fact that English is syntactically less complicated
than Swedish.
Furthermore, the text in Talbanken are from various brochures, newspapers and text books,
where idiomatic expressions are more likely the be appear and the language
presumably less strict than in Wall Street Journal\footnote{www.wsj.com}.
Also, the Penn Treebank contains a lower number of tags, 82 compared
to more than 300 in Talbanken. Even if the tags describing the same word
class as another tag are excluded, Talbanken still leaves us with more than 130 tags.
With more tags, we get more information, but as the number 
increase, so does the amount of work of finding the correct translation 
for each combination of tags and writing rules that cover all constructions.

@@

It has earlier been hard to identify missing constructions of the Swedish
implementation, since there was no large resource available to evaluate it on.
Our evaluations are based on Talbanken, and when first conducting tests,
we found much room for improvement.
the implementation of the languages needs to be more independent.
The resource grammar gives a good start and our present grammar 
covers constructions such as declarative sentences, questions, passives,
imperatives, relative clauses, cleft constructions etc.
During our work the grammar has been extended to cover language specific
constructions such as topicalisation and the features listed below. We further
put effort on avoiding and removing overgeneration, which might pass unnoticed
when parsing controlled text but which cause an explosion of ambiguous parse trees
when used for free text and with a large lexicon.
\subsection{The s-passive}
Passive voice is commonly used in Swedish.
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using the modal auxiliary verb `bli'
(\emph{`become'}) and the \textbf{s-passive} which is formed by adding an
\emph{s} to the verb: \\
\enumsentence{
\shortex{5}{Uppsatsen & skrev\textbf{s} & av & en & student.}
        {\emph{the essay} & \emph{wrote+\textbf{s}} & \emph{by} & \emph{a} & \emph{student}}
        {``The essay was written by a student."}}\label{sent:skrevs}
Some studies suggest that the s-passive is used in more than 80 \% of the times
\cite{laanemets}.
It is however not as common in the other Scandinavian languages,
and the resource grammar for Scandinavian therefore only implemented a function for
passive by using auxiliary verb.

@@

\enumsentence{Han s{\aa}g sina barns skor.\\ \label{ex:ReflGen}
He saw {\sc self's} children's shoes.}
\enumsentence{Sina vantar hade han gl{\"o}mt p{\aa} t{\aa}get.\\
{\sc self's} gloves, he had forgotten on the train.}
\enumsentence{Hon ber alla sina kompisar att g{\aa}\\
She asks all {\sc self's} friends to leave}\label{ex:ReflPredet}
\enumsentence{Jag vill att han rakar sig.\\
I want him to shave {\sc self}.}
\eenumsentence{
\item Hon tyckte om skolan och alla sina elever.\\
     She liked the school and all {\sc self's} students.
\item Han s{\aa}g sina f{\aa} b{\"o}cker och sin penna.\\
 He saw {\sc self's} few books and {\sc self's} pencil.
}\label{ex:ReflConj} \label{ex:ReflConj2}
Reflexive pronouns can not be used in subject noun phrases of finite sentences,
as shown by the ungrammatical examples in 
sentence (\ref{sent:sesig}) and (\ref{ex:ReflSubj}).
The third person reflexives (`sig',`sin') requires a third person antecedent (see \ref{ex:refljag}).
Furthermore, the antecedent must be within the same finite sentence as the reflexive
pronoun, see (\ref{sent:sesig}).
The grammar does not accept any of these sentences:
\enumsentence{
*Sina vantar var kvar p{\aa} t{\aa}get.\\
{\sc self's} gloves were left on the train.}\label{ex:ReflSubj}
\enumsentence{*Han och sin kompis läser en bok.\\
He and {\sc self's} friend are reading a book. }
\enumsentence{*Jag ger sina pengar till sina syskon.\\
I give {\sc self's} money to {\sc self's} siblings. }\label{ex:refljag}
\enumsentence{*Han vill att jag ser p{\aa} sig.\\
He wants me to look at {\sc self}.}\label{sent:sesig}

In order to achieve a scalable implementation of reflexive pronouns, we needed
to diverge from the old abstract implementation. To capture the agreement
between the antecedent and the reflexive pronoun, while still allowing the grammar
to accept reflexive constructions embedded in adverbial and adjectival phrases,
in fronted positions etc. we make use of the dependent types of GF. This is the
first use of the dependent types in the resource grammars and a could be
extended to other languages having similar dependencies.

\subsection{A second future tense: ``Kommer att"}
Swedish use two auxiliary verbs for future tense \\
\cite[p. 246]{H&H}:
\eenumsentence{
\item {
Det \textbf{kommer att} bli m{\"o}rkt snart\\
``It will get dark soon" }
\item {
Jag \textbf{ska} g{\aa} och l{\"a}gga mig nu\\
``I'm going to bed now"\\}
}
The resource grammar included 
\emph{`ska'}, which was implemented as the standard way of forming future
tense, and hence represents the translation of
\emph{`will'}. 
The new grammar also supports \emph{``kommer att"}, expressed as an alternative
future tense.
\begin{figure}[h]
\begin{verbatim}
(UseCl (TTAnt TFutKommer ASimul) PPos 
       (ImpersCl (AdvVP (ComplVA become_VA 
                (PositA dark_A)) soon_Adv))
\end{verbatim}
\caption{Result of parsing ``Det kommer att bli m{\"o}rkt snart". The future tense
         is marked by the constant TFutKommer}
  \label{fig:kommeratt}
\end{figure}

@@

Parsing bigger parts of Talbanken would hence give very low results at this stage, 
and a comparison of the results would not be of much value as
there would not be enough material
to do be able to do any interesting analysis.

@@

The result will be evaluated both automatically -- by comparing the output of
the translated trees from Talbanken -- and manually by professor Elisabet Engdahl.
She also evaluates the intermediate results.

@@

The usage of GF allows us to start from
a well-defined system for describing grammar, as well as tools for
parsing.
